A Tale of Two Transects

### 2019-11-07 writing/transect_paper.Rmd
sampling methods

@@

  - Brendan's pipeline:
      - Ion Torrent
          - used demultiplexing from SciLife (standard IonXpress)
          - primers removed with `cutadapt` [@martin2011]
      - PacBio
          - CCS basecalls re-done with `ccs` version ___ [@pacificbiosciences2019]
          - demultiplexed and primers removed with `cutadapt` [@martin2011]
      - Illumina MiSeq
          - pending
      - Region extraction with LSUx (available as R package on github and conda)
          - 5.8S located with Rfam [@kalvari2018] covariance model (CM) using `cmsearch` from Infernal 1.1.2 [@nawrocki2013]
          - read trimmed at 5' end of 5.8S; remaining read aligned to modified 5.8S + LSU (=32S) CM from Ribosomal Data Project (RDP) [@glockner2017, @cole2014] with annotations for different LSU domains (definitions from @michot1984 and @raue1988), using `cmalign` from Infernal 1.1.2
          - 5.8S, ITS2, D1, D2, D3, and conserved LSU domains (called LSU1, LSU2, LSU3, LSU4) extracted from alignment using annotations; ITS1 is everything before 5.8S.
      - extracted regions independently quality filtered by max expected error using `dada2::filterAndTrim()`
      - dereplicated using `dada2::derepFastq()`
      - denoised with `dada2::dada()`
      - de novo chimera removal using `dada2::removeBimeraDenovo()`
      - full length amplicons reconstructed from ASVs of extracted domains (r package tzara -- available on github and conda)
          - when a read is mapped to ASVs for all regions, they are concatenated.
          - bimera detection on each trio of variable-conserved-variable domains (e.g., ITS1-5.8S-ITS2, ITS2-LSU1-D1, etc.); reads which are detected at bimera for any trio are discarded.
          - For each ITS2 ASV, check each other domain:
              - If some reads are mapped to ASVs, but some are not, assign the unassigned to the closest ASV as long as distance < 10. This is less stringent than DADA2.
              - If no reads are mapped to ASVs, calculate consensus sequence:
                  - align reads using `DECIPHER` with RNA secondary structure calculation [@wright2015;@wright2016]
                  - remove outliers (i.e. chimeras) with `odseq` [@jehl2015]
                  - remove gap columns (>50%)
                  - take consensus (>50%) at non-gaps
                  - remove sequences with more than 3 ambiguous bases (bad consensus)
          - repeat process using each other domain as reference, in decreasing order of diversity.

      - taxonomy assignment
          - three methods: 
              - IDTAXA [@murali2018]
              - SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016]
              - RDP Naïve Bayesian Classifier (NBC) [@wang2007] as implemented in DADA2
          - three databases:
              - UNITE version 8 [@nilsson2019a]
                  - extracted ITS1 and ITS2 using ITSx [@Nilsson2010; @bengtsson-palme2013]
                  - trained three classifiers: full (for ITS, 5_8S, full amplicons), ITS1, ITS2.
              - RDP fungal LSU training set 11.5 [@cole2014]
                  - chose only sequences containing conserved primer sites for ITS4 (3' side of H11; CAUAUnAnUnAGnSSAGGA) and LR5 (on H37; CGAAnUUUCnCUCAGGAUAGCnG).
                  - ambiguities based on [@cannone2002; @tedersoo2015]
                  - Trim all bases after LR5 because they are not present in our amplicons.
                  - used to identify long amplicon, 32S, LSU
                  - taxonomy for non-fungi looked up from NCBI based on accession numbers
              - RDP Warcup fungal ITS training set [@wang2007]
                  - ITS1 and ITS2 extracted as for Unite;
          - taxonomic classification in the three datasets were standardized to @tedersoo2017c
              - full classification to genus level available in supplementary material
              - apparently used by Unite for nonfungi; @tedersoo2018 used by Unite for fungi
              - groups monophyletic
              - SINTAX and NBC require a ranked system, as opposed to more accepted classifications like @adl2019
      - ecological guild assigned using FUNGuild [@nguyen2016funguild]
      - alignment
          - preliminary alignment of 32S to RDP covariance model with `cmalign` from Infernal (aligns only conserved regions)
          - concatenate ITS1 to beginning of alignment
          - create guide tree using alignment of only conserved, gap-free columns in preliminary alignment, using UPGMA in DECIPHER [@wright2016]
          - realign nonconserved sites using MLocARNA (progressive simultaneous alignment and folding) from LocARNA 2.0.0RC8 [@will2007; @will2012; @will2013].

>>>>

Sampling was conducted at two sites approximately 30 km apart in the _Forêt Reservée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both locations were located in woodlands dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25 soil samples were collected at intervals of 1 m.
For each sample, any coarse organic debris was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 50 mg total of soil was collected from two locations in the homogenized soil and placed in Xpedition lysis solution (Zymo Research) and lysed in the field using a handheld bead-beater (Terralyzer; Zymo Research).
The remainder of each sample was dried at 40°C with an electric dehydrator within 24-48 hours of collection for bulk chemical analysis.

@@

  - Brendan's pipeline:
      - Ion Torrent
          - used demultiplexing from SciLife (standard IonXpress)
          - primers removed with `cutadapt` [@martin2011]
      - PacBio
          - CCS basecalls re-done with `ccs` version ___ [@pacificbiosciences2019]
          - demultiplexed and primers removed with `cutadapt` [@martin2011]
      - Illumina MiSeq
          - pending
      - Region extraction with LSUx (available as R package on github and conda)
          - 5.8S located with Rfam [@kalvari2018] covariance model (CM) using `cmsearch` from Infernal 1.1.2 [@nawrocki2013]
          - read trimmed at 5' end of 5.8S; remaining read aligned to modified 5.8S + LSU (=32S) CM from Ribosomal Data Project (RDP) [@glockner2017, @cole2014] with annotations for different LSU domains (definitions from @michot1984 and @raue1988), using `cmalign` from Infernal 1.1.2
          - 5.8S, ITS2, D1, D2, D3, and conserved LSU domains (called LSU1, LSU2, LSU3, LSU4) extracted from alignment using annotations; ITS1 is everything before 5.8S.
      - extracted regions independently quality filtered by max expected error using `dada2::filterAndTrim()`
      - dereplicated using `dada2::derepFastq()`
      - denoised with `dada2::dada()`
      - de novo chimera removal using `dada2::removeBimeraDenovo()`
      - full length amplicons reconstructed from ASVs of extracted domains (r package tzara -- available on github and conda)
          - when a read is mapped to ASVs for all regions, they are concatenated.
          - bimera detection on each trio of variable-conserved-variable domains (e.g., ITS1-5.8S-ITS2, ITS2-LSU1-D1, etc.); reads which are detected at bimera for any trio are discarded.

      - taxonomy assignment
          - three methods: 
              - IDTAXA [@murali2018]
              - SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016]
              - RDP Naïve Bayesian Classifier (NBC) [@wang2007] as implemented in DADA2
          - three databases:
              - UNITE version 8 [@nilsson2019a]
                  - extracted ITS1 and ITS2 using ITSx [@Nilsson2010; @bengtsson-palme2013]
                  - trained three classifiers: full (for ITS, 5_8S, full amplicons), ITS1, ITS2.
              - RDP fungal LSU training set 11.5 [@cole2014]
                  - used to identify long amplicon, LSU
                  - taxonomy for non-fungi looked up from NCBI based on accession numbers
              - RDP Warcup fungal ITS training set [@wang2007]
                  - ITS1 and ITS2 extracted as for Unite;
          - taxonomic classification in the three datasets were standardized to @tedersoo2017c
              - full classification to genus level available in supplementary material
              - apparently used by Unite for nonfungi; @tedersoo2018 used by Unite for fungi
              - groups monophyletic
              - SINTAX and NBC require a ranked system, as opposed to more accepted classifications like @adl2019
      - ecological guild assigned using FUNGuild [@nguyen2016funguild]
      - alignment as RNA in DECIPHER (iterative progressive alignment with RNS secondary structure prediction) [@wright2015]
      - tree using RAxML 8.2 [@stamatakis2014]; GTR+GAMMA; rapid bootstrapping with MRE-IGN stopping criterion [@pattengale2010].

<!--
          - preliminary alignment of 32S to RDP covariance model with `cmalign` from Infernal (aligns only conserved regions)
          - concatenate ITS1 to beginning of alignment
          - create guide tree using alignment of only conserved, gap-free columns in preliminary alignment, using UPGMA in DECIPHER [@wright2016]
          - realign nonconserved sites using MLocARNA (progressive simultaneous alignment and folding) from LocARNA 2.0.0RC8 [@will2007; @will2012; @will2013]. -->
